### 2017-11-16 manuscript.rmd
made zapta edits, pdf not rebuilt

^3^ Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA, 20095, USA

@@

KMRR [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture. The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event). The ortholog conjecture can and has been applied to diverse attributes, including molecular sequence, biochemical function, and, as in the study considered here, expression. It has important biological and technical implications. It shapes our understanding of the functional diversity of gene families. It is also used to relate findings from well-studied genes to homologous genes that have not been investigated in detail. It has been applied to many traits of genes, from gene sequence to biochemical properties to expression. While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.

@@

We next simulated the evolution of Tau under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation. The pairwise results of this heterogeneous model (Fig. 2*E*) are nearly indistinguishable from the results under the null model (Fig. 2*C*), and also have a higher correlation coefficient for orthologs (`r signif( ortholog_r_sim_oc$estimate, 2 )`, $P`r format_p( ortholog_r_sim_oc )`$) than paralogs (`r signif( paralog_r_sim_oc$estimate, 2 )`, $P`r format_p( paralog_r_sim_oc )`$). The phylogenetic analysis of the ortholog conjecture simulation (Fig. 2*F*) does reject the null hypothesis (Wilcoxon $P=`r signif( wilcox_oc( nodes_sim_oc_contrast ), 2 )`$).

@@

Another challenge is parameterizing differential expression in such a way that they can be analyzed in a phylogenetic context. Read counts generated by RNA-seq expression studies cannot be directly compared across species [@Dunn:2013cpa]. This is because there are various species-specific technical factors that can be mistaken for differences in expression across species. These can be canceled out within species before making comparisons across species [@Dunn:2013cpa].

@@

#### Phylogenetic signal and models

@@

We next approximated the Phylogenetic Independent Contrasts for the the genes that have $\Delta$AIC less than the threshold for OU under an OU model. These analyses do not reject the null hypothesis (Wilcoxon $P=`r  wilcox_ou %>% round( 3 )`$). The challenge with such an analysis is that BM is intrinsic to the classic implementation of Phylogenetic Independent Contrasts, which use the fact that the expected variance is proportional to branch length to scale each contrast. Under OU, branch length no longer has a linear relationship to expected variance. We therefore took an alternative simulation-based approach to scaling the OU contrasts. We implemented this new method, which we call picx (for "pic extended models"), in our R package hutan (https://github.com/caseywdunn/hutan). For each tree, picx first estimate the model parameters from the observed data. It then simulates two hundred datasets under these parameters, recording both the internal and tip node states. It then scales the observed contrasts by the mean magnitude of the corresponding contrasts in the simulations. We validated the method by comparing picx BM results to the standard pic results calculated by ape. These validations indicated a strong linear relationship between the two ($R^2$ > 0.95). This approach could be extended to approximate Phylogenetic Independent Contrasts under other continuous models of evolution as well.

@@

KMRR [@KryuchkovaMostacci:2016iw] Fig. 3 differs from their other analyses in that it presents independent changes in expression between triplets of genes. Each triplet has two paralogs that arose from a duplication event and one unduplicated homolog. They find that there is a tendency for the paralog with the lowest expression to have the highest Tau. From this they conclude that following duplication there is a common trajectory of evolutionary change, where one paralog evolves to have lower expression and also become more tissue specific. There is, however, a negative relationship between Tau and maximum expression across all genes (*SI Appendix* Fig. S6, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $p=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$). A simpler explanation for their plot is that it reflects the global tendency for Tau to decrease with maximum expression. The null expectation for any two genes sampled at random is for one to have higher maximum expression and lower Tau, and the other to have lower maximum expression and higher Tau. This global pattern could be due to both biological factors and the technical details of how Tau is calculated.

>>>>

^3^ Department of Ecology and Evolutionary Biology, University of California Los Angeles, Los Angeles, CA, 90095, USA

@@

KMRR [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture. The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event). The ortholog conjecture can and has been applied to diverse attributes, including molecular sequence, biochemical function, and, as in the study considered here, expression. It has important biological and technical implications. It shapes our understanding of the functional diversity of gene families. It is also used to relate findings from well-studied genes to homologous genes that have not been investigated in detail. While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.

@@

We next simulated the evolution of Tau under the ortholog conjecture, where the rate of evolution of Tau following duplication was `r dup_adjust` fold the rate following speciation. The pairwise results of this heterogeneous model (Fig. 2*E*) are nearly indistinguishable from the results under the null model (Fig. 2*C*), and also have a higher correlation coefficient for orthologs (`r signif( ortholog_r_sim_oc$estimate, 2 )`, $P`r format_p( ortholog_r_sim_oc )`$) than paralogs (`r signif( paralog_r_sim_oc$estimate, 2 )`, $P`r format_p( paralog_r_sim_oc )`$). The phylogenetic analysis of the ortholog conjecture simulation (Fig. 2*F*) does reject the null hypothesis (Wilcoxon $P=`r signif( wilcox_oc( nodes_sim_oc_contrast ), 2 )`$), as expected if the phylogenetic methods are working correctly.

@@

Another challenge is parameterizing differential expression in such a way that they can be analyzed in a phylogenetic context. Read counts generated by RNA-seq expression studies cannot be directly compared across species [@Dunn:2013cpa]. This is because there are various species-specific technical factors that can be mistaken for differences in expression across species. These can be canceled out within species before making comparisons across species.

@@

#### Phylogenetic signal and models of evolution

@@

We next approximated the Phylogenetic Independent Contrasts for the the genes that have $\Delta$AIC less than the threshold for OU under an OU model. These analyses do not reject the null hypothesis (Wilcoxon $P=`r  wilcox_ou %>% round( 3 )`$). The challenge with such an analysis is that BM is intrinsic to the classic implementation of Phylogenetic Independent Contrasts, which use the fact that the expected variance is proportional to branch length to scale each contrast. Under OU, branch length no longer has a linear relationship to expected variance. We therefore took an alternative simulation-based approach to scaling the OU contrasts. We implemented this new method, which we call picx (for "pic extended models"), in our R package hutan (https://github.com/caseywdunn/hutan). For each tree, picx first estimate the model parameters from the observed data. It then simulates two hundred datasets under these parameters, recording both the internal and tip node states. It then scales the observed contrasts by the mean magnitude of the corresponding contrasts in the simulations. We validated the method by comparing picx BM results to the standard pic results calculated by ape [@Paradis:2004dv]. These validations indicated a strong linear relationship between the two ($R^2=0.98$, $P=1.39\times 10^{−23}$). This approach could be extended to approximate Phylogenetic Independent Contrasts under other continuous models of evolution as well.

@@

KMRR [@KryuchkovaMostacci:2016iw] Fig. 3 differs from their other analyses in that it presents independent changes in expression between triplets of genes. Each triplet has two paralogs that arose from a duplication event and one unduplicated homolog. They find that there is a tendency for the paralog with the lowest expression to have the highest Tau. From this they conclude that following duplication there is a common trajectory of evolutionary change, where one paralog evolves to have lower expression and also become more tissue specific. There is, however, a negative relationship between Tau and maximum expression across all genes (*SI Appendix* Fig. S6, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $P=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$). A simpler explanation for their plot is that it reflects the global tendency for Tau to decrease with maximum expression. The null expectation for any two genes sampled at random is for one to have higher maximum expression and lower Tau, and the other to have lower maximum expression and higher Tau. This global pattern could be due to both biological factors and the technical details of how Tau is calculated.
